---
title: "How Do Socio-Economic Factors Influence Global Life Expectancy?"
author: "An Exploratory Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Project Overview

This project explores the relationship between global life expectancy and socio-economic indicators from 2000 to 2019. Using World Bank data, I focus on understanding how a country's wealth, education, and health investments relate to population health outcomes.

**Research Question:**

How have life expectancy trends evolved globally over the last two decades, and which socio-economic factors are most strongly associated with these trends?


## Dataset

**Source:** [Kaggle – "Life Expectancy & Socio-Economic Factors" (World Bank)](https://www.kaggle.com/datasets/mjshri23/life-expectancy-and-socio-economic-world-bank)

**All Relevant Variables:**

| Variable                           | Description                                                                                                                       |
| ---------------------------------- | --------------------------------------------------------------------------------------------------------------------------------- |
| **Country.Name**                   | Name of the country.                                                                                                              |
| **Country.Code**                   | country code.                                                                                                        |
| **Region**                         | Geographical region classification.                                                                                               |
| **IncomeGroup**                    | I ncome classification (Low, Lower Middle, Upper Middle, High).                                                         |
| **Year**                           | Observation year (2000–2019).                                                                                                     |
| **Life.Expectancy.World.Bank**     | Average life expectancy at birth, reported by the World Bank.                                                                     |
| **Health.Expenditure..**           | Current health expenditure as a percentage of GDP. Represents spending on healthcare services and goods consumed within the year. |
| **Education.Expenditure..**        | Total government expenditure on education as a percentage of GDP. Includes current and capital spending.                          |
| **Prevelance.of.Undernourishment** | Percentage of the population not consuming enough calories for an active, healthy life.                                           |
| **CO2**                            | CO2 emissions per capita (metric tons). A general indicator of industrialization and economic activity.                           |
| **Unemployment**                   | Percentage of the total labor force that is unemployed but actively seeking work.                                                 |
| **Corruption**                     | CPIA transparency and accountability rating, measuring public sector governance quality.                                          |
| **Sanitation**                     | Percentage of the population using safely managed sanitation services.                                                            |
| **Injuries**                       | Disability-Adjusted Life Years (DALYs) lost due to injuries. Higher values indicate worse injury-related health outcomes.         |
| **Communicable**                   | DALYs from communicable diseases. Reflects the burden of infectious disease.                                 |
| **NonCommunicable**                | DALYs from noncommunicable diseases. Higher values indicate greater chronic illness burden.         |


**Data Cleaning**

* Load dataset
```{r}
# Load dataset
raw_data <- read.csv("life expectancy.csv", header = TRUE)

# Make a copy of the dataset for cleaning
clean_data <- raw_data
```

* Standardize column names
```{r}
names(clean_data) <- make.names(names(clean_data), unique = TRUE)
```

* Remove duplicates and empty strings
```{r}
clean_data <- clean_data |> distinct()

clean_data <- clean_data |>
mutate(across(where(is.character), ~na_if(., ""))) |>
mutate(across(where(is.character), trimws))
```

* Convert numeric-looking strings to numeric
```{r}
clean_data <- clean_data |>
mutate(across(where(function(x) all(grepl("^[0-9.]+$", x[!is.na(x)]))), as.numeric))
```

* Filter years 2000–2019
```{r}
if ("Year" %in% names(clean_data)) {
clean_data <- clean_data |> filter(Year >= 2000, Year <= 2019)
}
```

* Drop rows with missing `Life.expectancy`
```{r}
if ("Life.expectancy" %in% names(clean_data)) {
clean_data <- clean_data |> drop_na(Life.expectancy)
}
```

* Save cleaned dataset as `cleaned_life_expectancy.csv`
```{r}
write.csv(clean_data, "cleaned_life_expectancy.csv", row.names = FALSE)
cat("File saved: cleaned_life_expectancy.csv\n")
```


## Exploring the Data

Get the dimensions (rows and columns)
```{r}
dim(clean_data)
```

View first few rows
```{r}
head(clean_data)
```

View column names
```{r}
colnames(clean_data)
```

View data structures and types
```{r}
str(clean_data)
```


## Descriptive Statistics
```{r}
numeric_cols <- clean_data |> select(where(is.numeric))
summary(numeric_cols)
```

## Variable Selection

Although the dataset contains additional variables (CO2 emissions, unemployment rates, disease spread, corruption ratings), this analysis focuses on the variables most directly tied to life expectancy:

1. health spending
2. education spending
3. income level

Other variables are listed as potentially related but not explored in depth due to the project scope.


## Potential Relationships We Expect
1. Life expectancy increases over time
2. Higher health expenditure likely -> higher life expectancy
3. Higher education expenditure -> higher life expectancy
4. Higher income group -> higher life expectancy


### 1. How Has Global Life Expectancy Evolved Over Time?

* **Visualization:** Line plot of average life expectancy between 2000 and 2019.
```{r}
clean_data |>
  group_by(Year) |>
  summarize(mean_life = mean(Life.Expectancy.World.Bank, na.rm = TRUE)) |>
  ggplot(aes(x = Year, y = mean_life)) +
  geom_line(linewidth = 1.2, color = "palegreen3") +
  labs(
    title = "Global Average Life Expectancy (2000–2019)",
    x = "Year",
    y = "Average Life Expectancy"
  ) +
  theme_minimal()
```

* **Insight:** Average global life expectancy shows a steady incline from 2000 to 2019. Although the rate of improvement slows slightly in later years, the overall rise suggests continuous improvement in public health, medical access, and living conditions across most regions.


### 2. How Does Health Expenditure Influence Life Expectancy?

* **Visualization:** Scatterplot with regression line.
```{r}
clean_data |>
  ggplot(aes(x = Health.Expenditure.., y = Life.Expectancy.World.Bank)) +
  geom_point(alpha = 0.4, color = "lightblue2") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "lightblue3") +
  labs(
    title = "Health Expenditure vs. Life Expectancy",
    x = "Health Expenditure (% of GDP)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()

```

* **Insight:** Countries with higher health expenditure tend to have higher life expectancy. The positive regression line implies that wealthier nations generally provide better healthcare, nutrition, and infrastructure. However, the scatter plot includes numerous outliers so we cannot say for sure that there is a direct correlation without considering other factors.


### 3. How Does Education Expenditure Influence Life Expectancy?

* **Visualization:** Scatterplot with regression line.
```{r}
clean_data |>
  ggplot(aes(x = Education.Expenditure.., 
             y = Life.Expectancy.World.Bank)) +
  geom_point(alpha = 0.4, color = "cornflowerblue") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "royalblue4") +
  labs(
    title = "Education Expenditure vs. Life Expectancy",
    x = "Education Expenditure (% of GDP)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()
```

* **Insight:** Higher education expenditure is associated with higher life expectancy, but the relationship is typically weaker and more variable than health expenditure. This makes sense because investing in education indirectly supports health by improving health literacy, economic opportunities, and long-term outcomes, but the effect is not as direct as healthcare investments.


### 4. How Does Life Expectancy Differ By Income Group?

* **Visualization:** Boxplot grouped by income classification.
```{r}
clean_data$IncomeGroup <- factor(
  clean_data$IncomeGroup,
  levels = c("Low income", "Lower middle income", "Upper middle income", "High income")
)

clean_data |>
  ggplot(aes(
    x = IncomeGroup,
    y = Life.Expectancy.World.Bank,
    fill = IncomeGroup
  )) +
  geom_boxplot(alpha = 0.8) +
  labs(
    title = "Life Expectancy by Income Group",
    x = "Income Group",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Spectral")

```

* **Insight:** Life expectancy increases clearly across income groups, with high-income countries showing the highest median values and low-income countries showing the lowest. The separation between groups highlights persistent global health inequalities tied to resources, infrastructure, and access to medical care. Variation within groups suggests that income level is important but not the only factor influencing population health.



## Data Limitations

* Dataset focuses on national-level data, not individual-level outcomes
* Regional differences within countries are not captured
* Some socio-economic indicators (ex. CO2 emissions, corruption) were omitted to simplify analysis
* Missing data and small variations in reporting may bias results



## Conclusion

The analysis shows that global life expectancy has steadily improved from 2000 to 2019, but not evenly across countries. Wealthier nations consistently achieve higher life expectancy, and both GDP and education spending show clear positive relationships with population health. When grouped by income level, the differences become even more clear because high-income countries live significantly longer than low-income countries. Overall, the findings suggest that strong economies, well-funded education systems, and broader social investments play key roles in supporting longer, healthier lives.



*GitHub Repository:* [https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23](https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23)


