---
title: "How Do Socio-Economic Factors Influence Global Life Expectancy?"
author: "An Exploratory Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

## Project Overview

This project explores the relationship between global life expectancy and socio-economic indicators from 2000 to 2019. Using World Bank data, I focus on understanding how a country's wealth, education, and health investments relate to population health outcomes.

**Research Question:**

How have life expectancy trends evolved globally over the last two decades, and which socio-economic factors are most strongly associated with these trends?


## Dataset

**Source:** [Kaggle – "Life Expectancy & Socio-Economic Factors" (World Bank)](https://www.kaggle.com/datasets/mjshri23/life-expectancy-and-socio-economic-world-bank)

**Features Selected for Analysis:**

| Variable              | Description                               |
| --------------------- | ----------------------------------------- |
| Year                  | Observation year (2000–2019)              |
| Life Expectancy       | Average life expectancy in years          |
| Health Expenditure    | % of GDP spent on healthcare              |
| Education Expenditure | % of GDP spent on education               |
| Income Group          | Income classification (Low, Middle, High) |
| Undernourishment      | % of population lacking adequate food     |

**Data Cleaning**

* Load dataset
```{r}
# Load dataset
raw_data <- read.csv("life expectancy.csv", header = TRUE)

# Make a copy of the dataset for cleaning
clean_data <- raw_data
```

* Standardize column names
```{r}
names(clean_data) <- make.names(names(clean_data), unique = TRUE)
```

* Remove duplicates and empty strings
```{r}
clean_data <- clean_data |> distinct()

clean_data <- clean_data |>
mutate(across(where(is.character), ~na_if(., ""))) |>
mutate(across(where(is.character), trimws))
```

* Convert numeric-looking strings to numeric
```{r}
clean_data <- clean_data |>
mutate(across(where(function(x) all(grepl("^[0-9.]+$", x[!is.na(x)]))), as.numeric))
```

* Filter years 2000–2019
```{r}
if ("Year" %in% names(clean_data)) {
clean_data <- clean_data |> filter(Year >= 2000, Year <= 2019)
}
```

* Drop rows with missing `Life.expectancy`
```{r}
if ("Life.expectancy" %in% names(clean_data)) {
clean_data <- clean_data |> drop_na(Life.expectancy)
}
```

* Save cleaned dataset as `cleaned_life_expectancy.csv`
```{r}
write.csv(clean_data, "cleaned_life_expectancy.csv", row.names = FALSE)
cat("File saved: cleaned_life_expectancy.csv\n")
```


## Exploring the Data

Get the dimensions (rows and columns)
```{r}
dim(clean_data)
```

View first few rows
```{r}
head(clean_data)
```

View column names
```{r}
colnames(clean_data)
```

View data structures and types
```{r}
str(clean_data)
```


## Potential Relationships We Expect
1. Higher health expenditure likely → higher life expectancy
2. Higher education expenditure → higher life expectancy
3. Higher income group → higher life expectancy
4. Higher undernourishment → lower life expectancy


## Descriptive Statistics
```{r}
numeric_cols <- clean_data |> select(where(is.numeric))
summary(numeric_cols)
```


## Exploratory Data Analysis

## Anova of Health expenditure & life expectancy 
anova_health_life_expenditure <- aov(cleaned_life_expectancy$Health.Expenditure ~ cleaned_life_expectancy$Life.Expectancy.World.Bank, data = df)
summary(anova_health_life_expenditure) 

##Anova of education expenditure & life expectancy 
anova_education_life_expenditure <- aov(cleaned_life_expectancy$Education.Expenditure ~ cleaned_life_expectancy$Life.Expectancy.World.Bank, data = df) 
summary(anova_education_life_expenditure) 

##Anova of income group expenditure & life expectancy 
anova_income_life_expenditure <- aov(Life.Expectancy.World.Bank ~ IncomeGroup,data = cleaned_life_expectancy) 
summary(anova_income_life_expenditure) 

##Anova of undernourishment & life expectancy  
anova_undernourishment_life_expenditure <- aov(Life.Expectancy.World.Bank ~ Prevelance.of.Undernourishment,data = cleaned_life_expectancy)  
summary(anova_undernourishment_life_expenditure)



### 1. Global Life Expectancy Trends

* **Visualization:** Line plot of average life expectancy over time.
```{r}
clean_data |>
  group_by(Year) |>
  summarize(mean_life = mean(Life.Expectancy.World.Bank, na.rm = TRUE)) |>
  ggplot(aes(x = Year, y = mean_life)) +
  geom_line(linewidth = 1.2, color = "palegreen3") +
  labs(
    title = "Global Average Life Expectancy (2000–2019)",
    x = "Year",
    y = "Average Life Expectancy"
  ) +
  theme_minimal()
```
* **Insight:** 

### 2. Relationship Between Health Expenditure and Life Expectancy

* **Visualization:** Scatterplot with regression line.
```{r}
clean_data |>
  ggplot(aes(x = Health.Expenditure.., y = Life.Expectancy.World.Bank)) +
  geom_point(alpha = 0.4, color = "lightblue2") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "lightblue3") +
  labs(
    title = "Health Expentiture vs. Life Expectancy",
    x = "Health Expentiture (% of GDP)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()

```
* **Insight:** 

### 3. Relationship Between Education Expenditure and Life Expectancy

* **Visualization:** Scatterplot with regression line.
```{r}
clean_data |>
  ggplot(aes(x = Education.Expenditure.., 
             y = Life.Expectancy.World.Bank)) +
  geom_point(alpha = 0.4, color = "cornflowerblue") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "royalblue4") +
  labs(
    title = "Education Expenditure vs. Life Expectancy",
    x = "Education Expenditure (% of GDP)",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal()
```
* **Insight:** 

### 4. Life Expectancy by Income Group

* **Visualization:** Boxplot grouped by income classification.
```{r}
clean_data |>
  ggplot(aes(x = IncomeGroup, 
             y = Life.Expectancy.World.Bank,
             fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.8, palete = "") +
  labs(
    title = "Life Expectancy by Income Group",
    x = "Income Group",
    y = "Life Expectancy (Years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
* **Insight:** 


## Data Limitations

* Dataset focuses on national-level data, not individual-level outcomes
* Regional differences within countries are not captured
* Some socio-economic indicators (ex. CO₂ emissions, corruption) were omitted to simplify analysis
* Missing data and small variations in reporting may bias results



## Conclusion

This analysis highlights that life expectancy trends are strongly associated with health, education, and income group. While these indicators explain much of the variation, other unmeasured factors likely play a role. Policy makers can use these insights to prioritize investments in health and education to improve life expectancy outcomes globally.



*GitHub Repository:* [https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23](https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23)

