---
title: "How do Socio-Economic Factors Influence Global Life Expectancy?"
author: "An Exploratory Data Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview

This project explores the relationship between global life expectancy and socio-economic indicators from 2000 to 2019. Using World Bank data, I focus on understanding how a country's wealth, education, and health investments relate to population health outcomes.

**Key Question:**

* How have life expectancy trends evolved globally over the last two decades, and which socio-economic factors are most strongly associated with these trends?


## Dataset

**Source:** Kaggle – "Life Expectancy & Socio-Economic Factors" (World Bank)
**Features Selected for Analysis:**

| Variable              | Description                               |
| --------------------- | ----------------------------------------- |
| Year                  | Observation year (2000–2019)              |
| Life.expectancy       | Average life expectancy in years          |
| GDP                   | Gross Domestic Product per capita         |
| Education.expenditure | % of GDP spent on education               |
| Health.expenditure    | % of GDP spent on healthcare              |
| Income.group          | Income classification (Low, Middle, High) |
| Undernourishment      | % of population lacking adequate food     |

**Data Cleaning**

* Load dataset
```{r}
# Load dataset
raw_data <- read.csv("life expectancy.csv", header = TRUE)

# Make a copy of the dataset for cleaning
clean_data <- raw_data
```

* Standardize column names
```{r}
names(clean_data) <- make.names(names(clean_data), unique = TRUE)
```

* Remove duplicates and empty strings
```{r}
clean_data <- clean_data %>% distinct()

clean_data <- clean_data %>%
mutate(across(where(is.character), ~na_if(., ""))) %>%
mutate(across(where(is.character), trimws))
```

* Convert numeric-looking strings to numeric
```{r}
clean_data <- clean_data %>%
mutate(across(where(function(x) all(grepl("^[0-9.]+$", x[!is.na(x)]))), as.numeric))
```

* Filter years 2000–2019
```{r}
if ("Year" %in% names(clean_data)) {
clean_data <- clean_data %>% filter(Year >= 2000, Year <= 2019)
}
```

* Drop rows with missing `Life.expectancy`
```{r}
if ("Life.expectancy" %in% names(clean_data)) {
clean_data <- clean_data %>% drop_na(Life.expectancy)
}
```

* Save cleaned dataset as `cleaned_life_expectancy.csv`
```{r}
write.csv(clean_data, "cleaned_life_expectancy.csv", row.names = FALSE)
cat("File saved: cleaned_life_expectancy.csv\n")
```



## Exploratory Data Analysis

### 1. Global Life Expectancy Trends

* **Visualization:** Line plot of average life expectancy over time.
* **Insight:** 

### 2. Relationship Between GDP and Life Expectancy

* **Visualization:** Scatterplot with regression line.
* **Insight:** 

### 3. Relationship Between Education Expenditure and Life Expectancy

* **Visualization:** Scatterplot with regression line.
* **Insight:** 

### 4. Life Expectancy by Income Group

* **Visualization:** Boxplot grouped by income classification.
* **Insight:** 



## Potential Relationships




## Simplified Descriptive Statistics




## Data Limitations

* Dataset focuses on national-level data, not individual-level outcomes
* Regional differences within countries are not captured
* Some socio-economic indicators (ex. CO₂ emissions, corruption) were omitted to simplify analysis
* Missing data and small variations in reporting may bias results



## Conclusion

This analysis highlights that life expectancy trends are strongly associated with GDP, education, and income group. While these indicators explain much of the variation, other unmeasured factors likely play a role. Policy makers can use these insights to prioritize investments in health and education to improve life expectancy outcomes globally.



*GitHub Repository:* [https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23](https://github.com/SNazari23/INST447_Group_Project/tree/mainSNazari23)
